Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-01-10 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Removing second chunk
And adding a paragrah with a very short resume.

% -- Start spin-off -------------------
\section{Enhancing {\sf NodIO} with a Web Worker-based architecture}
\label{sec:w2}

In order to improve the number of cycles per user, we enhanced the
basic architecture in several different aspects.  One improvement 
involved restarting the client once the solution was
found, so that as long as the user was visiting the page it was
running an evolutionary algorithm. Another was a small algorithmic
enhancement: instead of having a fixed population,
population size was randomly distributed between 128 and 256 and
different for each client.  The other improvement was to 
use the Web Worker API as described next.

JavaScript has been commonly used to develop client-side scripts
that handle the user interface, communicating asynchronously with servers and
updating the content that is displayed \cite{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the
responsiveness of the web application, because everything is handled by a
single thread. The Web Workers specification \cite{hickson2012web} defines an
API that allows Web developers to execute long-running scripts in parallel
with the scripts of the main page. Worker scripts are spawned in the
background allowing a thread-like operation using message-passing as the
communication mechanism between the worker's execution environment and the
main thread. According to the Web Workers W3C Candidate Recommendation
\cite{hickson2012web} workers are expected to be long-lived, they have a high
start-up performance cost, and a high per-instance memory cost. The Web Worker
API is implemented in current versions of both desktop and mobile web browsers.
For volunteer computing using the Web Worker API brings several advantages
over a single threaded implementation:

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even in case the user brings forward another tabbed
window.
\item Several evolutionary algorithms can be executed in parallel in a single web
page. 
\item Many implementations of the Web Worker API can use multiple-core CPUs for
the execution of worker scripts, in particular the execution in multiple-core CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each {\sf NodEO} instance can be restarted independently.
\end{itemize}

A sequence diagram of the enhanced NodIO, which is called NodIO-W$^2$, is presented
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two {\sf NodEO} instances per browser. The sequence diagram shows the
interaction between processes in a time sequence. In this version there are
two kinds of asynchronous messages: HTTP Requests shown as bold arrows between
the client and the server where PUT and GET HTTP methods are represented by
red and blue arrows respectively. There are also messages between workers and
the main thread locally in the client, which are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em Node.js} This is the server side process which is basically
  the same as used previously, except for some minor log-related
  changes. 
\item {\em Client main thread} The main script creates
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronous HTTP requests to the server are implemented using the
XMLHttpRequest object.  
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}
\label{fig:system:w2}
\end{figure}
%
% End spin-off ----------------------------------------------------------------
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}

% - Start spin-off again
% -------------------------------------------------------

The sequence of events happens as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in Node.js. As part of the
initialization procedure, the sequential number of experiments is initialized and a log
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP
request, then the page is loaded.
\item The main script renders the page and creates the worker instance. In
order to create a dedicated worker instance, another GET request is sent to the
server, then the client receives the second worker implementation script.
Each worker instance executes over time one or more evolutionary algorithms
{\em islands}. Each island starts with a random population and it is assigned a
universally unique identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.
The initial parameters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following the previously
specified parameters. Once the population has been initialized, the EA
generation loop is started. After {\em n} generations the state of the population
is evaluated. The chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not the global best,
the EA loop must continue, but first a chromosome is randomly selected from
the shared population. The current generation number and the best
chromosome are also sent to the client's main thread in order to update the DOM to
display the current state of the island (described next). These messages are
all asynchronous, so the EA loop of
the worker continues even before a message is received by the target. If the best fitness is a global best, the best
chromosome is sent to the server and the loop ends. A message is sent to the
main thread indicating that the best individual has been found.
\item The main thread receives messages from workers as events handled by
a callback function. When an iteration message is received, the page is
updated by displaying a dynamic plot of the generation number and best
fitness, and a representation of the chromosome is also generated.
\item When a global best is received from an island, the current experiment
ends, the experiment number is incremented, and the population array is reset.
\item When the main thread receives a message indicating that a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the start-up cost, only the parameters and population are reset and a new UUID
assigned.
\end{enumerate}
% End spin-off--------------------------------------------

\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the shades of gray the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}

\subsection{First experiments}
\label{sec:w2Exp}

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table*}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%

>>>>

\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the shades of gray the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}

\subsection{Experiments using web workers}
\label{sec:w2Exp}

The results reported above indicated that we should look for a way of
increasing the time that every user devotes. This led us to create an
architecture that uses Web Workers, which is described in
\cite{2016arXiv160101607M}; web workers work in the background even if
the web page is not in focus. A few improvements were also tested,
basically using a random population and restarting the web worker when
the solution was found.

The experiments were run in the same way as before, using several
 announcements in Twitter throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table*}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of {\tt PUT}s per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.83719  &  25.962  $\pm$ 1.22015 & 1.242   $\pm$
0.04746 \\
Weibull & ND & 75.11457519 $\pm$ 2.97470251  & 0.69714410 $\pm$ 0.01383976 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
